Parsing Postfix log files

### 2007-09-07 doc/logparser.tex
Improved stats: * Automated extracting data from directories of results; it's not quite right,   as making changes to the directories or contents doesn't cause a rebuild. * Split normal/shuffle/reverse factor graph into two, so they're legible. * Updated log 22 stats, and graphs. * Generate stats for logs 22 and 62-68 mean and standard deviation.
Added introductions and conclusions to several of the sections. Moved some subsections because it made more sense. Some minor improvements. Automated more stats generation. Made 'make veryclean' remove everything (or nearly everything) that's generate

    Parsing Postfix logs is much more difficult than it first appears but
    it is possible to achieve a high degree of accuracy in understanding
    the logs, and thus reconstructing the actions taken by Postfix to
    generate the logs.  This paper describes the creation of a parser,
    documenting the parsing algorithm and rules, explaining the
    difficulties encountered and the solutions developed, with reference to
    an implementation which stores data gleaned from the logs in an SQL
    database.  The gathered data can then be used to optimise current
    anti-spam measures, provide a baseline to test new anti-spam measures
    against, or to produce statistics showing how effective those measures
    are.

@@

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table.} will check
whether the IP address of the connected client matches the left hand side
of each line in the file and will return the right hand side of the first
matching line.  Other restrictions determine their result in different
ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks whether the
IP address of the client is present in the listed RBL\@.

@@

\subsection{Parser background}

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules are solely concerned with
identifying a line and extracting data from it, whereas the algorithm's
task is to follow the journey each mail takes through Postfix, piecing the
data extracted by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.

@@

@@

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in section~\ref{rules} and the fields in particular in
section~\ref{rule attributes}, and that discussion won't be repeated here.
The data saved from parsing the logs is divided into two tables described
below: connections and results.

Every connection and/or mail will have a single entry in the connections
table containing the following fields:

Every log line corresponding to Postfix performing an action other than
INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
SMTP command, delivering a mail, or bouncing a mail.  Any log line where
the Postfix action is INFO is not interesting in and of itself, but
provides additional information which will be saved with other results;
IGNORED log lines don't even provide useful information.

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows new rules to be written and tested without requiring modifications
to the algorithm source code (significantly lowering the barrier to entry
for new or casual users), and greatly simplifies both algorithm and rules.
It also creates a clear separation of functionality: rules handle low level
details of identifying log lines and extracting data from a line, whereas
the algorithm handles the higher level details of following the path a mail
takes through Postfix, assembling the required data, etc.

\subsection{Rule attributes}

@@

@@

@@

    \item [shuffled] 6--11\% increase; mean 8.34, standard deviation 2.44.

    \item [reversed] 15--22\% increase; mean 17.97, standard deviation
        2.94.

Overall this increases the parser's efficiency by 5.5--10\%; a modest but
worthwhile increase.

@@

Perl compiles the original regex into an internal representation,
optimising the regex to improve the speed of matching, but this compilation
and optimisation takes CPU time.  Perl automatically caches static regexs,
but dynamic regexs need to be explicitly compiled and cached.
Graph~\ref{normal regex vs discard regex} shows execution times with and
without caching the regex; also shown on the graph are the file sizes and
number of lines in each file.  Caching the compiled regexs is obviously far
more efficient; graph~\ref{normal regex vs disarded regex factor} shows the
percentage execution time increase when not caching each regex is between
350--570\% (mean 465.58\%, standard deviation 114.70\%).

@@

Where the rules are quite simple and each rule is completely independent of
its fellows, the algorithm is significantly more complicated and highly
internally interdependent.  The algorithm deals with all the complications
of parsing, the eccentricities and oddities of Postfix logs, and presents
the resulting data in a normalised, easy to use representation.

\subsection{A naive approach}

@@

Unfortunately it's not that straightforward.

@@

This flow chart shows the paths the data representing a mail/connection can
take through the parser algorithm.

@@

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The data flow chart in section~\ref{flow-chart}
should also be consulted while reading this section.

@@

@@

@@

@@

@@

\subsubsection{Percentage increase of shuffled + reversed over normal}
\label{normal vs shuffled vs reversed ordering factor}
\includegraphics{plot-normal-shuffle-reverse-factor.ps}

@@

\begin{tabular}[]{lrr}
                                    & Mean      & Standard Deviation    \\
    All logs                        & 465.58\%  & 114.70\%              \\
    All logs except 22 and 62--68   & 497.43\%  & 50.58\%               \\
    Logs 22 and 62--68              & 127.16\%  & 21.28\%               \\

>>>>

    Parsing Postfix logs is much more difficult than it first appears, but
    it is possible to achieve a high degree of accuracy in understanding
    the logs, and thus reconstructing the actions taken by Postfix to
    generate the logs.  This paper describes the creation of a parser,
    documenting the parsing algorithm and rules, explaining the
    difficulties encountered and the solutions developed, with reference to
    an implementation which stores data gleaned from the logs in an SQL
    database.  The gathered data can then be used to optimise current
    anti-spam measures, provide a baseline to test new anti-spam measures
    against, or to produce statistics showing how effective those measures
    are.

@@

\subsection{Introduction}

This section provides background information helpful in understanding the
remainder of the paper.  It begins with a discussion of the motivation
underlying the project, followed by some technical information: the use of
a database as an API\@; a brief introduction to the SMTP protocol; and a
longer introduction to Postfix, concentrating on the topics most relevant
to this paper, namely Postfix restrictions and policy servers.  The
assumptions made in designing and implementing the parser are explained, as
are the conventions used in this paper.  Other projects which attempt to
parse Postfix logs are summarised (full details are available in
Section~\ref{other-parsers}), plus a review of previously published
research in this area.

XXX SHOULD THERE BE A REGEX SECTION\@?

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table.  XXX PROVIDE A
REFERENCE} will check
whether the IP address of the connected client matches the left hand side
of each line in the file and will return the right hand side of the first
matching line.  Other restrictions determine their result in different
ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks whether the
IP address of the client is present in the listed RBL\@.

@@

@@

\subsection{Conclusion}

This section has provided background information on several topics relevant
to the remainder of the paper.  It started with the motivation behind the
project, continuing with explanations of:

    \item Using a database as an application programming interface.

    \item The SMTP protocol.

    \item Postfix restrictions and policy servers.

The assumptions which must be satisfied for the parser to work correctly
were discussed; the conventions used in the paper were listed; a brief
comparison of this project against other Postfix parsers was provided,
followed by a review of the previously published literature.

@@

The database is an integral part of the parser: it stores the rules and the
results of applying those rules to Postfix log files.  Understanding the
database schema is important in understanding the actions of the parser,
and essential to developing further applications which utilise the data
gathered.

\subsection{Introduction}

The database schema can be conceptually divided in two: the rules which are
used to parse log files, and the data saved from the parsing of log files.
Rules have the fields required to parse the log lines, extract data to be
saved, and the action to be executed; they also have several fields which
aid the user in understanding what each rule parses.  The rules are
described in detail in section~\ref{rules} but the fields are covered in
section~\ref{rule attributes}.

The data saved from parsing the logs is also divided into two tables as
described below: connections and results.  The connections table contains a
row for every mail accepted and every connection where there was a
rejection; the individual fields will be described in the forthcoming
section~\ref{connections table}.  The results table has one or more rows
per row in the connections table, depending on the specifics of the
mail/connection; the fields will be covered in detail in
section~\ref{results table}.

\subsection{Rules table}

@@

Rules are discussed in detain in section~\ref{rules}, but the rules table
is covered here.  Rules are created by the user, not the parser, and will
not be modified by the parser (except for the hits and hits\_total fields).
Rules parse the individual lines, telling the parser which fields to
extract and what action to take for that line.

@@

\label{connections table}

Every accepted mail and every connection where there was a rejection will
have a single entry in the connections table containing the following
fields:

\label{results table}

Every log line corresponding to Postfix performing an action other than
INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
SMTP command, delivering a mail, or bouncing a mail.  Any log line where
the Postfix action is INFO is not interesting in and of itself, but
provides additional information which will be saved with other results;
IGNORED log lines don't even provide useful information.  Each row is
associated with a single connection, though there may be many results per
connection.

\subsection{Conclusion}

The table containing the rules used by the parser, and the two tables
containing the data extracted from the Postfix logs were described, with
the purpose of each field discussed in detail.  A clear, comprehensible
schema is essential in using the extracted data; it's more important for
using the data than storing the data, because storing the data is a
write-once operation, whereas utilising the data requires frequent
searching, sorting and manipulation of the data to produce customised
reports and/or statistics.

\subsection{Introduction}

This section discusses the rules used in parsing Postfix log files,
starting with rule characteristics, continuing with detecting and dealing
with overlapping rules, and the problems they can cause.  An example rule
and a line it would match are provided, plus a description of how the
fields in the rule are used when matching and performing the action (the
details of the table containing the rules have already been described in
section~\ref{rule attributes},).  This section finishes with a discussion
of rule efficiency concerns, referring to the graphs in
section~\ref{graphs}.

\subsection{Rule characteristics}

The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  The complexity and variation in Postfix's
logs requires similar flexibility in the parser; decoupling the parsing
rules from the associated actions allows new rules to be written and tested
without requiring modifications to the algorithm source code (significantly
lowering the barrier to entry for new or casual users), and greatly
simplifies both algorithm and rules.  It also creates a clear separation of
functionality: rules handle low level details of identifying log lines and
extracting data from a line, whereas the algorithm handles the higher level
details of following the path a mail takes through Postfix, assembling the
required data, etc.

@@

\input{stats-normal-shuffle-reverse-percentage}

Overall this provides a modest but worthwhile performance increase.

@@

Perl compiles the original regex into an internal representation,
optimising the regex to improve the speed of matching, but this compilation
and optimisation takes CPU time.  Perl automatically caches static regexs,
but dynamic regexs need to be explicitly compiled and cached.
Graph~\ref{normal regex vs discard regex} shows execution times with and
without caching the regex; also shown on the graph are the file sizes and
number of lines in each file.  Caching the compiled regexs is obviously far
more efficient; graph~\ref{normal regex vs disarded regex factor} shows the
percentage execution time increase when not caching each regex.  Over all
93 log files the execution time increase when not caching each compiled
regex works out at:

XXX IMPROVE THIS\@.

\input{stats-cached-discarded-include-for-discussion}

\subsection{Conclusion}

This section dealt with the rules used in parsing Postfix log files:

    \item The characteristics of the rules were described.

    \item Detecting overlapping rules and dealing with the problems they
        can cause was covered, including a discussion of why overlapping
        rules can be helpful as well as harmful.

    \item An example log line and the rule matching it was given, followed
        by a description of how the fields in the rule are used both in the
        matching and the subsequent action that is executed.

    \item The database table containing the rules is dealt with in
        section~\ref{rule attributes}, and is not duplicated in this
        section.

    \item The final part of this section discusses rule efficiency,
        covering the effects of caching compiled regexs and optimal
        ordering ordering of rules, with reference to the graphs in
        section~\ref{graphs}.

@@

Where the rules are quite simple and each rule is completely independent of
its fellows, the algorithm is significantly more complicated and highly
internally interdependent.  The algorithm deals with all the complications
of parsing, the eccentricities and oddities of Postfix logs, and presents
the resulting data in a normalised, easy to use representation.  The
algorithm's task is to follow the journey each mail takes through Postfix,
piecing the data extracted by rules into a coherent whole, saving it in a
useful and consistent form, and performing housekeeping duties.

\subsection{Introduction}

This section covers the following topics:

    \item A high level overview of the algorithm.

    \item The first set of complications encountered: initially obvious
        difficulties which had to be overcome.

    \item A flow chart showing the paths a mail can take through the
        algorithm.

    \item A description of the full algorithm.

    \item The actions which can be taken by the parser are described in
        detail.

    \item Additional complications which have arisen during the development
        of this parser are described.

\subsection{A high level overview}

@@

Unfortunately that ignores the many complications encountered.

@@

This flow chart shows the paths the data representing a mail/connection can
take through the parser algorithm.  The flow chart covers the most common
paths only; there are additional, uncommon paths which are excluded for the
sake of clarity.

@@

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The data flow chart in section~\ref{flow-chart}
should also be consulted while reading this section.  For the sake of
clarity this description covers the most common paths through the
algorithm, as including every path would hinder understanding.  The more
uncommon paths are caused by the complications described in
section~\ref{additional complications}, and are covered therein.

@@

XXX EXPLAIN EXACTLY WHERE IN THE ALGORITHM THE \newline CHECKS ETC\@. KICK
IN\@.

@@

\subsection{Conclusion}

@@

\subsection{Introduction}

@@

\subsection{Introduction}

\subsection{Parsers reviewed}

@@

\subsubsection{Percentage increase of shuffled + reversed over normal}
\label{normal vs shuffled vs reversed ordering factor}
\includegraphics{plot-normal-shuffle-factor.ps}
\newline
\includegraphics{plot-normal-reverse-factor.ps}

@@

\input{stats-cached-discarded-include-for-graph}

\subsection{Conclusion}
